Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Added annotation erasure operation.

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
annotations $T$ in recursion $\fix[\ell] T t$.  The type constructor $T$
has to be monotone in the size argument; this is a technical condition
for type-based termination \cite{gimenez:typeBased}.  We will make use
of it in Section~\ref{sec:comp}.

@@

@@

In the following, we list syntactic properties of our judgements.  To
this end, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

All types in a context are considered in the resurrected context,
which justifies the first statement of the following lemma.
A resurrected context is more permissive, as it brings more variable
into scope.  As such, it is comparable to an extended context or a
context where types have been replaced by subtypes.  This intuition
accounts for the remaining statements but (\ref{it:ressubst}).  The
latter is a defining property of substitutions: only replacement for
irrelevant sizes may refer to irrelevant size variables.
\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der \sigma : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der \sigma : \Delta$.
\item \label{it:ressubst} If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

The relation $\sueq \Gamma \sigma {\sigma'} \tau \Delta$ is a partial equivalence
relation (PER) on term-side substitutions $\sigma,\sigma'$. Note that
we cannot resurrect this judgement to
$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau \Delta$ in general!
The type-side substitution $\tau$ can be converted to one that is
equal in and for resurrected contexts.
\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}

>>>>

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
We have \fbox{boxed} the rules dealing with irrelevant size application.
Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
annotations $T$ in recursion $\fix[\ell] T t$.  The type constructor $T$
has to be monotone in the size argument; this is a technical condition
for type-based termination \cite{gimenez:typeBased}.  We will make use
of it in Section~\ref{sec:comp}.

@@

Typing requires from annotations $\ann a$ in a term only that they are well-scoped size expressions, \ie, just mention relevant size variables.  Let \fbox{$t^\infty$} denote the \emph{erasure} of term $t$, meaning that we replace all annotations $\ann a$ in $t$ by $\ann \infty$.  Let \fbox{$t \approx u$} relate terms that only differ in their annotations, \ie, $t \approx u \defiff t^\infty = u^\infty$.
Erasure does not change the term modulo judgmental equality:
\begin{lemma}[Erasure and similarity]
\label{lem:erase}
\bla
\begin{enumerate}
\item If\/ $\Gamma \der t : T$ then $\Gamma \der t = t^\infty : T$.
\item If\/ $\Gamma \der t, u : T$ and $t \approx u$ then $\Gamma \der t = u : T$.
\end{enumerate}
\end{lemma}

@@

\noindent
In the following, we list syntactic properties of our judgements.  To
this end, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

All types in a context are considered in the resurrected context,
which justifies the first statement of the following lemma.
A resurrected context is more permissive, as it brings more variable
into scope.  As such, it is comparable to an extended context or a
context where types have been replaced by subtypes.  This intuition
accounts for the remaining statements but (\ref{it:ressubst}).  The
latter is a defining property of substitutions: only replacement for
irrelevant sizes may refer to irrelevant size variables.
\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$.
  Then $\resurrect \Gamma \der \sid : \Gamma$, which can be written
  $\sid : \resurrect\Gamma \leq \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der \sigma : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der \sigma : \Delta$.
\item \label{it:ressubst} If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

The relation $\sueq \Gamma \sigma {\sigma'} \tau \Delta$ is a partial equivalence
relation (PER) on term-side substitutions $\sigma,\sigma'$. Note that
usually we cannot resurrect this judgement to
$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau {\resurrect\Delta}$.
For instance,
$\sueq{z_1 {\div} \Size.\,z_2 {\div} \Size} {\sg{z_1}} {\sg{z_2}} {\sg{\infty}} {z {\div} \Size}$
holds but
$\sueq{z_1 {:} \Size.\,z_2 {:} \Size} {\sg{z_1}} {\sg{z_2}} {\sg{\infty}} {z {:} \Size}$
clearly not.
% The type-side substitution $\tau$ can be converted to one that is
% equal in and for resurrected contexts.
\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}
